---
title: "Getting started with censusapi"
output: 
  rmarkdown::html_vignette:
    df_print: kable
pkgdown:
  as_is: true
description: >
  If you're new to censusapi, start here.
vignette: >
  %\VignetteIndexEntry{Getting started with censusapi}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(message = FALSE, 
											warning = FALSE,
											purl = NOT_CRAN,
											comment = "#>")
```

`censusapi` is a wrapper for the United States Census Bureau's [APIs](https://www.census.gov/data/developers/data-sets.html). As of 2017 over 200 Census API endpoints are available, including Decennial Census, American Community Survey, Poverty Statistics, and Population Estimates APIs. This package is designed to let you get data from all of those APIs using the same main function—`getCensus`—and the same syntax for each dataset.

`censusapi` generally uses the APIs' original parameter names so that users can easily transition between Census's documentation and examples and this package. It also includes metadata functions to return data frames of available APIs, variables, and geographies.

This package returns the data as-is with the original variable names created by the Census Bureau and any quirks inherent in the data. Each dataset is a little different. Some are documented thoroughly, others have documentation that is sparse. Sometimes variable names change between years or geographies are added or dropped. This package can't overcome those challenges, but tries to make it easier to get the data for use in your analysis. Make sure to thoroughly read the [documentation](https://www.census.gov/data/developers/data-sets.html) for your dataset and see below for how to [get help with Census data.](#Other-ways-to-get-help) 

## API key setup
To use the Census APIs, [sign up](http://api.census.gov/data/key_signup.html) for an API key, which will be sent to your provided email address. You'll need that key to use this package. `censusapi` will use it by default without any extra work on your part. 

Within R, run:
```{r, eval = FALSE}
# Add key to .Renviron
Sys.setenv(CENSUS_KEY=YOURKEYHERE)
# Reload .Renviron
readRenviron("~/.Renviron")
# Check to see that the expected key is output in your R console
Sys.getenv("CENSUS_KEY")
```

Once you've added your census key to your system environment, censusapi will use it by default without any extra work on your part. 

In some instances you might not want to put your key in your .Renviron - for example, if you're on a shared school computer. You can always choose to manually set `key = "YOURKEY"` as an argument in `getCensus()` if you prefer.

## Finding your API
To get started, load the `censusapi` library.
```{r}
library(censusapi)
```
The Census APIs have over 200 endpoints, covering dozens of different datasets.

To see a current table of every available endpoint, run `listCensusApis`:
```{r, eval = FALSE}
apis <- listCensusApis()
View(apis)
```

This returns useful information about each endpoint, including `name`, which you'll need to make your API call.

## Using `getCensus`
The main function in `censusapi` is `getCensus`, which makes an API call to a given Census API and returns a data frame of results. Each API has slightly different parameters, but there are always a few required arguments:

* `name`: the name of the API as defined by the Census, like "acs5" or "timeseries/bds/firms"
* `vintage`: the dataset year, generally required for non-timeseries APIs
* `vars`: the list of variable names to get
* `region`: the geography level to return, like state or county

Some APIs have additional required or optional arguments, like `time`, `monthly`, or `period`. Check the specific [documentation](https://www.census.gov/data/developers/data-sets.html) for your API to see what options are allowed.

Let's walk through an example getting uninsured rates by income group using the [Small Area Health Insurance Estimates API](https://www.census.gov/data/developers/data-sets/Health-Insurance-Statistics.html), which provides detailed annual state-level and county-level estimates of health insurance rates.

### Choosing variables
`censusapi` includes a metadata function called `listCensusMetadata` to get information about an API's variable options and geography options.
Let's see what variables are available in the SAHIE API:

```{r}
sahie_vars <- listCensusMetadata(
	name = "timeseries/healthins/sahie", 
	type = "variables")
head(sahie_vars)
```
We'll use a few of these variables to get uninsured rates by income group:

* `IPRCAT`: Income Poverty Ratio Category
* `IPR_DESC`: Income Poverty Ratio Category Description
* `PCTUI_PT`: Percent Uninsured in Demographic Group for Selected Income Range, Estimate
* `NAME`: Name of the geography returned (e.g. state or county name)

### Choosing regions
We can also use `listCensusMetadata` to see which geographic levels we can get data for using the SAHIE API.
```{r}
listCensusMetadata(
	name = "timeseries/healthins/sahie", 
	type = "geography")
```
This API has three geographic levels: `us`, `county` within states, and `state`.

First, using `getCensus`, let's get uninsured rate by income group at the national level for 2018.
```{r}
sahie_national <- getCensus(
	name = "timeseries/healthins/sahie",
	vars = c("NAME", "IPRCAT", "IPR_DESC", "PCTUI_PT"), 
	region = "us:*", 
	year = 2019)
sahie_national
```

We can also get this data at the state level for every state by changing `region` to `"state:*"`:
```{r}
sahie_states <- getCensus(
	name = "timeseries/healthins/sahie",
	vars = c("NAME", "IPRCAT", "IPR_DESC", "PCTUI_PT"), 
	region = "state:*", 
	year = 2019)
head(sahie_states)
```

Finally, we can get county-level data. The geography metadata showed that we can choose to get county-level data within states. We'll use `region` to specify county-level results and `regionin` to request data for Alabama and Alaska.
```{r}
sahie_counties <- getCensus(
	name = "timeseries/healthins/sahie",
	vars = c("NAME", "IPRCAT", "IPR_DESC", "PCTUI_PT"), 
	region = "county:*", 
	regionin = "state:01,02", 
	year = 2019)
head(sahie_counties, n=12L)
```

Because the SAHIE API is a timeseries (as indicated in its name), we can get multiple years of data at once using the `year` argument.
```{r}
sahie_years <- getCensus(
	name = "timeseries/healthins/sahie",
	vars = c("NAME", "PCTUI_PT"), 
	region = "state:01,17", 
	year = "2006:2019")
```

Here's a simple chart of that yearly data using the [ggplot2](https://ggplot2.tidyverse.org/) package.
```{r}
library(ggplot2)
ggplot(sahie_years, aes(x = YEAR, y = PCTUI_PT, group = state, color = state)) +
	geom_line() +
	scale_y_continuous(
		breaks = seq(6, 18, 2),
		limits = c(6, 18),
		expand = c(0, 0)) +
	theme(
		legend.position = "top",
		panel.grid.minor = element_blank(),
		panel.grid.major.x = element_blank(),
		panel.border = element_blank(),
		axis.title = element_blank()) +
	scale_color_discrete(
		labels = c("Alabama", "Illinois"), 
		breaks = c("01", "17")) +
	labs(
		title = "Percent of people who are uninsured by state",
		subtitle = "For people below age 65",
		caption = "Source: Small Area Health Insurance Estimates"
	)
```

### Annotations
Some Census datasets, inclduing the American Community Survey, use annotated values. These values use numbers or symbols to indicate that the data is unavailable, has been topcoded, has an insufficient sample size, or other noteworthy characteristics. Read more from the Census Bureau on [ACS annotation meanings](https://www.census.gov/data/developers/data-sets/acs-1year/notes-on-acs-estimate-and-annotation-values.html) and [ACS variable types](https://www.census.gov/data/developers/data-sets/acs-1year/notes-on-acs-api-variable-types.html).

The `censusapi` package is intended to return the data as-is so that you can receive those unaltered annotations. If you are using data for a small geography make sure to check for values like `-666666666` or check the annotation columns for non-empty values to exclude as needed.

As an example, we'll get median income with associated annotations and margin of error, for three census tracts in Washington, DC — one that is available, one that is topcoded, and one that is unavailable. Notice that income is topcoded at $250,000 — meaning any tract's income that is above that threshold is listed as $250,000. Some are not available and are listed as -666666666. You can see these annotations in the "EA" (estimate annotation) and "MA" (margin of error annotation) columns.


```{r}
acs_income <- getCensus(
	name = "acs/acs5",
	vintage = 2020, 
	vars = c("B19013_001E", "B19013_001EA", "B19013_001M", "B19013_001MA"), 
	region = "tract:006804,007703,000903",
	regionin = "county:001&state:11")
acs_income
```

### Variable groups
For some surveys, particularly the American Community Survey and Decennial Census, you can get many related variables at once using a `group`, defined by the Census Bureau. In some other data tools, like American FactFinder, this idea is referred to as a `table`.

The American Community Survey (ACS) APIs include estimates (variable names ending in "E"), annotations, margins of error, and statistical significance, depending on the data set. Read more on [ACS variable types](https://www.census.gov/data/developers/data-sets/acs-5year.html) and annotation symbol meanings on the Census website. 

You can retrieve these annotation variables manually, by specifying a list of variables. We'll get the estimate, margin of error and annotations for median household income in the past 12 months for Census tracts in Alaska.
```{r}
acs_income <- getCensus(
	name = "acs/acs5",
	vintage = 2017, 
	vars = c("NAME", "B19013_001E", "B19013_001EA", "B19013_001M", "B19013_001MA"), 
	region = "tract:*", 
	regionin = "state:02")
head(acs_income)
```

You can also retrieve also estimates and annotations for a group of variables in one command. Here's the `group` call for that same table, [B19013](https://api.census.gov/data/2016/acs/acs5/groups/B19013.html).
```{r}
# See descriptions of the variables in group B19013
group_B19013 <- listCensusMetadata(
	name = "acs/acs5",
	vintage = 2017,
	type = "variables",
	group = "B19013")
group_B19013

acs_income_group <- getCensus(
	name = "acs/acs5", 
	vintage = 2017, 
	vars = c("NAME", "group(B19013)"), 
	region = "tract:*", 
	regionin = "state:02")
head(acs_income_group)
```


### Advanced geographies

Some geographies, particularly Census tracts and blocks, need to be specified within larger geographies like states and counties. This varies by API endpoint, so make sure to read the documentation for your specific API and run `listCensusMetadata` to see the available geographies.

You may want to get get data for many geographies that require a parent geography. For example, tract-level data from the 2010 Decennial Census can only be requested from one state at a time. 

In this example, we use the built in `fips` list of state [FIPS codes](https://www.census.gov/library/reference/code-lists/ansi.html) to request tract-level data from each state and join into a single data frame.
```{r}
tracts <- NULL
for (f in fips) {
	stateget <- paste("state:", f, sep="")
	temp <- getCensus(
		name = "dec/sf1",
		vintage = 2010,
		vars = "P001001",
		region = "tract:*",
		regionin = stateget)
	tracts <- rbind(tracts, temp)
}
head(tracts)
```

The `regionin` argument of `getCensus` can also be used with a string of nested geographies, as shown below.

The 2010 Decennial Census summary file 1 requires you to specify a state and county to retrieve block-level data. Use `region` to request block level data, and `regionin` to specify the desired state and county.
```{r}
data2010 <- getCensus(
	name = "dec/sf1",
	vintage = 2010,
	vars = "P001001", 
	region = "block:*",
	regionin = "state:36+county:027+tract:010000")
head(data2010)
```

For many more examples and advanced topics check out [all of the articles](https://www.hrecht.com/censusapi/articles/.)

## Troubleshooting
The APIs contain hundreds of API endpoints and dozens of datasets, each of which work a little differently. The Census Bureau also makes frequent updates, which unfortunately are not always announced in advance. If you're getting an error message or unexpected results, here are some things to check.

### Variables
Use `listCensusMetadata(type = "variables")` on your API to see the table of available variables.

* Occasionally the variable names will change with data updates or API updates. The names may be different from year to year. Capitalization may also be inconsistent between datasets or years.
* The Census APIs are case-sensitive, which means that if the variable name you want is uppercase you'll need to write it uppercase in your request. Most of the APIs use uppercase variable names, but some use lowercase and some even use sentence case. 

### Geographies
Use `listCensusMetadata(type = "geographies")` on your dataset to check which geographies you can use.

* Each API has its own list of valid geographies and they occasionally change as the Census Bureau makes updates. If a previously available geography isn't available anymore, email cnmp.developers.list@census.gov detailing the issue.
* If you're specifying a region by FIPS code, for example `state:01`, make sure to use the full code, padded with 0s if necessary. The APIs did not always enforce this (previously, `state:1` usually worked), but now they do. See the Census [reference files](https://www.census.gov/geographies/reference-files/2017/demo/popest/2017-fips.html) for valid FIPS codes.

### Unexpected errors
Occasionally you might get the general error message `"There was an error while running your query.  We've logged the error and we'll correct it ASAP.  Sorry for the inconvenience."` This comes from the Census Bureau and could be caused by any number of problems, including server issues. Try rerunning your API call. If that doesn't work and you are requesting a large amount of data, try reducing the amount that you're requesting, for example getting only one state at a time. If you're still having trouble, email census.data@census.gov. Include in your email the raw API call that's provided in your `getCensus` error message (not your R code) so that they can try to help.

### Other ways to get help
* Open a [Github issue](https://github.com/hrecht/censusapi/issues) for bugs or issues with this R package.
* Join the public Census Bureau [Slack channel](http://links.govdelivery.com/track?type=click&enid=ZWFzPTEmbWFpbGluZ2lkPTIwMTgwOTE3Ljk0OTc1NjkxJm1lc3NhZ2VpZD1NREItUFJELUJVTC0yMDE4MDkxNy45NDk3NTY5MSZkYXRhYmFzZWlkPTEwMDEmc2VyaWFsPTE3MjA3MjUwJmVtYWlsaWQ9aHJlY2h0MkBnbWFpbC5jb20mdXNlcmlkPWhyZWNodDJAZ21haWwuY29tJmZsPSZleHRyYT1NdWx0aXZhcmlhdGVJZD0mJiY=&&&105&&&https://join.slack.com/t/uscensusbureau/shared_invite/enQtMjQ3NzUyNTM3NDU3LTZmNGI1MmQzY2Y2ZTU1ODJhNDQwMmY2YmZiNmFkNzg4YmJkYmQzZjQyNDhkNDYxN2JhYjkxZDEwMGI2OGU5NzQ?eml=gd&utm_medium=email&utm_source=govdelivery) and ask your question in the R or API rooms.
* Email the Census Bureau API team at census.data@census.gov for questions relating to the underlying data and APIs.

## Additional resources
* [Census Data API User Guide](https://www.census.gov/data/developers/guidance/api-user-guide.Overview.html)

